---
title: "NYPD Shooting Incident Data"
author: "Cody Hill"
date: "2023-04-04"
output:
  pdf_document: default
  html_document: default
---

## Setup 

We will first begin by loading in the packages we intend to use. 

Then, importing the data using a URL directly from the source. 
This ensures we will capture updates to the data as they come in whenever we run this again.

```{r Setup RMD}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 12, fig.height = 8) 

library(tidyverse)
library(ggmap)
library(ggplot2)

import_URL <- read.csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
```

## Exploratory Data Analysis (EDA)
Let's take a look at the dimension of this imported data.frame as well as the variable types of each column.
```{r EDA}
dim(import_URL)
str(import_URL)
```

## Use GGMAP to Visualize Locations
Here we're going to visualize the location of each shooting incident using the coordinates given in the dataset,
and superimposing them over a background image of the area. We can use the minimum and maximum values 
of each coordinate to find our map's bounding box then use `ggmap()` to do the heavy lifting.
```{r Map Visualization}
# Initialize the bounding box that will contain the map coordinates.
map_bounds <- c(left = min(import_URL$Longitude), 
        bottom = min(import_URL$Latitude),
        right = max(import_URL$Longitude), 
        top = max(import_URL$Latitude))

# Plot of incident coordinates
incident_map_point <- ggmap(get_stamenmap(map_bounds, maptype = 'terrain', zoom = 11)) + 
        geom_point(data = import_URL, 
                aes(x = Longitude, y = Latitude),
                color = 'darkred', 
                size = 0.25, 
                alpha = 0.2) +
        ggtitle('Point Plot of NYPD Shooting Incident Reporting 2006 - 2021\n    Source:<https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic>') +
        labs(x = 'Longitude', y = 'Latitude')
incident_map_point

# Density Map to better visualize density of points.
incident_map_density <- ggmap(get_stamenmap(map_bounds, maptype = 'terrain', zoom = 11)) + 
        stat_density2d_filled(data = import_URL, contour_var = 'density',
                aes(x = Longitude, y = Latitude, fill = after_stat(level)), 
                bins = 20, 
                geom = 'polygon', 
                alpha = 0.8) +
        geom_density_2d(data = import_URL, 
                aes(x = Longitude, y = Latitude), 
                bins = 20, 
                alpha = 0.2, 
                color = "white") +
        guides(fill = guide_legend(title = "Density")) + 
        ggtitle('Density Plot of NYPD Shooting Incident Reporting 2006 - 2021\n    Source:<https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic>') +
        labs(x = 'Longitude', y = 'Latitude')
incident_map_density
```